---
title: "Culminating Activity"
author: "Aloba, Loyan Edrian | Dalmonte, Franklin | Delrio Martin Nicholas | Dorin, Homer Adriel | Rivera, Nythan"
date: ""

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Abstract
Provide a concise summary of the study including purpose, methods, results, and conclusions (150–250 words).

# Chapter 1: Introduction

## Background of the Study
The integration of Generative AI (GenAI) into software engineering has been aggressively marketed as an unprecedented paradigm shift in productivity. Industry narratives frequently promise that AI coding assistants will streamline workflows, accelerate output, and consequently elevate both financial compensation and developer well-being. However, emerging empirical research reveals a significant productivity trade-off inherent in GenAI adoption. Utilizing the SPACE framework for developer productivity, Afroz et al. (2025) found that while developers using AI tools experience localized spikes in activity and perceive themselves as operating faster, they do not necessarily produce better software or experience increased fulfillment. In fact, rigorous randomized controlled trials indicate that developers using early-2025 AI tools took 19% longer to complete complex tasks, despite subjectively believing they were working 20% faster—a cognitive disconnect establishing an industry-wide "productivity placebo" (Becker et al., 2025).
This paradox is driven by severe cognitive and operational friction. As AI accelerates initial code generation, it shifts systemic bottlenecks downstream, increasing pull request review times by 91% and inflating the volume of bugs per developer (Faros AI, 2025). Developers are forced to spend more time debugging hallucinated code and navigating complex, AI-generated architectural debt, which has been shown to actively damage macroscopic software delivery performance and system stability (DORA Research Program, 2024).
Consequently, this study investigates whether the financial rewards of AI adoption genuinely translate into higher job satisfaction, or if the persistent friction of modern coding creates a "GenAI Productivity-Satisfaction Tradeoff." In this paradigm, developers may be caught in a cycle where hyper-compensation—driven by a reported 56% AI skill premium (PricewaterhouseCoopers, 2025)—acts as a retention mechanism despite high stress, a concept previously identified in occupational psychology as job embeddedness or "golden handcuffs" (Mitchell et al., 2001). As AI increasingly shifts the developer's role from writing original code to reviewing and debugging AI output, the industry must question whether these tools actually improve the human experience of software engineering. Utilizing secondary data from the Stack Overflow 2024 Developer Survey (Stack Overflow, 2024), this study builds a predictive model to determine if AI tools genuinely improve job satisfaction, or if developers are simply compensated more to endure increased friction.


## Statement of the Problem
This study investigates the systemic impact of AI adoption on software developers. With Job Satisfaction as the primary target variable, the study asks:
To what extent does the use of AI development tools predict a developer’s overall job satisfaction?
How does workflow friction (time spent searching for answers and debugging) interact with AI usage to influence this satisfaction?
Does a higher salary offset the negative impact of workflow friction when predicting overall job satisfaction?


## Objectives of the Study
### General Objective
To investigate the "GenAI Productivity-Satisfaction Tradeoff" by developing a predictive model that evaluates how AI tool adoption, financial compensation, and workflow friction interact to determine a software developer's overall job satisfaction.

### Specific Objectives
To determine the extent to which AI tool adoption serves as a predictor for both financial compensation and workflow friction (time spent searching and debugging).

To identify the primary predictors of developer job satisfaction using a Classification and Regression Tree (CART) machine learning model.

To evaluate the interaction between compensation and workflow friction, specifically testing if a higher salary offsets the negative predictive impact of AI-induced friction on overall job satisfaction.


## Hypotheses (if applicable)
H0 (Null Hypothesis): AI usage, financial compensation, and workflow friction do not form a significant predictive model for developer job satisfaction, and no distinct tradeoff exists between these variables.

H1 (Alternative Hypothesis): The optimal predictive model for Job Satisfaction will demonstrate a significant tradeoff, where the positive predictive value of high financial compensation is canceled out by the negative predictive impact of AI-induced workflow friction.


## Significance of the Study
This research benefits technology companies, engineering managers, and developers. By identifying the gap between productivity metrics such as salary and output, and human metrics such as satisfaction, organizations can re-evaluate how they integrate AI. It shifts the focus from simply buying AI licenses to actively reducing the systemic friction that causes developer burnout.


## Scope and Limitations
The study analyzes secondary data from the Stack Overflow 2024 Developer Survey. The scope is strictly limited to respondents who identify as "Employed, full-time" to maintain a stable economic baseline.

A critical limitation is that survey data regarding compensation and satisfaction is highly susceptible to MNAR (Missing Not At Random) bias, as extreme earners and highly burned-out developers frequently skip these questions. To mitigate this, advanced algorithmic imputation using MICE via Classification and Regression Trees was utilized to predict and fill missing values rather than relying on listwise deletion.

This study does not attempt an exhaustive feature selection of the entire Stack Overflow dataset. Instead, the predictor variables (AI Usage, Salary, Time Spent Searching, Years of Experience, and Remote Status) were handpicked based on intuition and prior literature regarding developer productivity.  	


# Chapter 2: Review of Related Literature

## Related Studies
Summarize previous research.

## Related Literature


## Conceptual Framework
At the core of this framework is the ultimate target variable: Job Satisfaction.

Rather than utilizing isolated linear regressions, the framework employs a Decision Tree architecture where predictor variables interact organically:

Primary Predictor: AI Tool Adoption is positioned as the central independent variable, hypothesized to simultaneously impact financial compensation and workflow friction.

Interacting Predictors: Financial Compensation (Salary) and Workflow Friction (Time Spent Searching) act as mediating variables within the model.
Contextual Controls: Years of Experience and Remote Status serve as baseline variables to account for demographic and workplace variances.

The CART algorithm evaluates how these variables interact, specifically testing whether high compensation mathematically buffers the negative impact of AI-induced workflow friction when predicting final job satisfaction levels.

```{r conceptual-framework, echo=FALSE, fig.width=10, fig.height=5.8, out.width='100%', fig.cap="Conceptual Framework of the GenAI Productivity-Satisfaction Tradeoff"}
old_par <- par(mar = c(1, 1, 0.6, 1))

plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 8))
box(col = "#666666")
title(main = "Conceptual Framework of the GenAI Productivity-Satisfaction Tradeoff", cex.main = 1.0, line = 0.5)

draw_box <- function(xmin, ymin, xmax, ymax, label, col, border, cex = 0.88) {
  rect(xmin, ymin, xmax, ymax, col = col, border = border, lwd = 2)
Hyperparameter Shuffling (Model-Choice Robustness)
}

draw_arrow <- function(x0, y0, x1, y1, label = NULL, y_offset = 0) {
  arrows(x0, y0, x1, y1, length = 0.08, lwd = 1.8, col = "#666666")
  if (!is.null(label)) {
source("main.r")
shuffle <- read_csv("outputs/cart_hyperparam_shuffle.csv", show_col_types = FALSE)

best_row <- shuffle %>%
  filter(status == "ok") %>%
  arrange(xerror, desc(nsplit)) %>%
  slice(1)

best_row
```

This check keeps the imputed dataset fixed and varies cp, minsplit, and maxdepth to see how pruning and tree depth change the structure of the model. It is a model-choice robustness check, not a data-uncertainty check. The results are saved to [outputs/cart_hyperparam_shuffle.csv](outputs/cart_hyperparam_shuffle.csv).

In the shuffling, cp acts as the pruning strength and xerror is the cross-validated error from rpart. Higher cp values prune more aggressively and can remove all splits. This is visible in [outputs/cart_hyperparam_shuffle.csv](outputs/cart_hyperparam_shuffle.csv), where cp values of 0.008 to 0.02 yield nsplit = 0. The best-performing combination was cp = 0.001, minsplit = 20, and maxdepth = 5, with xerror = 0.9872315 and nsplit = 170. Under this setting, variable importance favored Years_Exp (1676.70) and Salary (872.31), with smaller contributions from Is_Remote (263.38) and Uses_AI (203.72). TimeSearching_Min did not appear in the main tree (importance = 0) for this specific imputation.

```{r}
# Placeholder: Hyperparameter shuffle xerror vs cp
```

This plot will show how pruning strength affects cross-validated error. A higher cp means stronger pruning and can collapse the tree into no splits. The lowest xerror with non-zero splits is the basis for selecting the final hyperparameters.

```{r}
# Placeholder: Variable importance bar chart for the main tree
```

The main-tree importance chart will highlight the primary drivers of job satisfaction under the best hyperparameter setting from the shuffling. It should emphasize that these importance values come from a single, fixed imputation and therefore reflect model-choice effects rather than imputation uncertainty.

Seed Sensitivity Check (Data Uncertainty)

```{r seed-sensitivity, eval=FALSE}
library(dplyr)
library(readr)

# Run the seed sensitivity check and load the summaries
source("cart_robustness_check.r")
seed_runs <- read_csv("outputs/cart_mice_seed_runs.csv", show_col_types = FALSE)
seed_averages <- read_csv("outputs/cart_mice_seed_averages.csv", show_col_types = FALSE)

seed_averages
```

This check fixes the model settings but re-imputes the missing data across several seeds. It uses seeds 42, 100, 200, 300, and 400 with m = 5 and maxit = 5, and it holds the CART settings constant at cp = 0.001, minsplit = 20, and maxdepth = 5. The full results are written to [outputs/cart_mice_seed_runs.csv](outputs/cart_mice_seed_runs.csv), and the per-seed averages are written to [outputs/cart_mice_seed_averages.csv](outputs/cart_mice_seed_averages.csv).

The seed-sensitivity check in [outputs/cart_mice_seed_averages.csv](outputs/cart_mice_seed_averages.csv) shows how results shift when the imputed data changes. Across seeds 42, 100, 200, 300, and 400, TimeSearching_Min becomes the dominant signal (range: 2444.99 to 2509.76). In the same runs, Years_Exp stays moderate (range: 970.66 to 1112.81) and Salary stays lower (range: 342.53 to 684.11). Uses_AI drops to near zero in most runs (range: 0 to 2.75), and Is_Remote stays very small (range: 1.23 to 4.56). The xerror values remain tightly clustered between 0.97838 and 0.97942, with split counts ranging from 159 to 192.

```{r}
# Placeholder: Seed sensitivity plot (importance by seed)
```

This plot will compare variable importance across imputation seeds to show how data uncertainty affects the model. Stable predictors will show consistent importance, while unstable predictors will fluctuate or drop out.

CP Profile (Pruning Check)

```{r cp-profile, eval=FALSE}
library(readr)

cp_profile <- read_csv("outputs/cart_cp_profile.csv", show_col_types = FALSE)
cp_profile
```

The CP profile is generated in [cart_robustness_check.r](cart_robustness_check.r) using a baseline imputation from the first seed. The profile table is saved to [outputs/cart_cp_profile.csv](outputs/cart_cp_profile.csv) and is used to explain how error changes as the tree is pruned. This gives a second, independent check on the pruning choice beyond the hyperparameter shuffling alone.
draw_arrow(6.1, 5.8, 7.2, 4.2, "Higher compensation may\nreduce the negative effect\nof friction", 0.12)
draw_arrow(6.1, 2.5, 7.2, 3.7, "Workflow friction may\nlower job satisfaction", -0.1)
draw_arrow(3.0, 6.8, 4.0, 6.1)
draw_arrow(8.2, 1.6, 8.1, 3.0)

par(old_par)
```

The diagram illustrates the interacting variables within the predictive CART model. AI Tool Adoption acts as the primary predictor, hypothesized to trigger a dual pathway: an increase in Financial Compensation and a simultaneous increase in Workflow Friction. These mediating variables converge on the target variable, Job Satisfaction, where the algorithm will test if the financial benefits successfully buffer the friction-induced burnout. In this statistical context, "buffering" refers to a moderating effect where high compensation offsets, mitigates, or increases a developer's tolerance for the negative psychological toll of AI-driven workflow friction (Lam, et al., 2025). Finally, Prior Years of Experience and Remote Work Status are integrated into the system as contextual controls that adjust the model's baseline predictions. Unlike a sequential linear regression, the CART algorithm evaluates how all of these variables split and compound simultaneously to determine the final satisfaction outcome.

## Chapter 3: Methodology
Research Design

This study employs a quantitative, descriptive-correlational research design. It utilizes multiple linear regression modeling to examine the predictive relationships between AI adoption, compensation, and workflow friction.

Data Collection


```{r}
# Load dataset
# data <- read.csv("your_data.csv")
# head(data)
```

Variables Description


Data Analysis Techniques


## Chapter 4: Results and Discussion
Descriptive Statistics

```{r descriptive-statistics}
library(dplyr)
library(readr)

df <- read_csv("outputs/data_imputed_primary.csv", show_col_types = FALSE)

summary_df <- df %>%
  summarise(
    n = n(),
    Job_Satisfaction_mean = mean(Job_Satisfaction, na.rm = TRUE),
    Job_Satisfaction_sd = sd(Job_Satisfaction, na.rm = TRUE),
    Salary_mean = mean(Salary, na.rm = TRUE),
    Salary_sd = sd(Salary, na.rm = TRUE),
    TimeSearching_Min_mean = mean(TimeSearching_Min, na.rm = TRUE),
    TimeSearching_Min_sd = sd(TimeSearching_Min, na.rm = TRUE),
    Years_Exp_mean = mean(Years_Exp, na.rm = TRUE),
    Years_Exp_sd = sd(Years_Exp, na.rm = TRUE),
    Uses_AI_rate = mean(as.numeric(as.character(Uses_AI)) == 1, na.rm = TRUE),
    Is_Remote_rate = mean(as.numeric(as.character(Is_Remote)) == 1, na.rm = TRUE)
  )

summary_df
```

Visualization
```{r}
# Placeholder for charts/plots
```


This section summarizes the sample used in the analysis, including central tendencies for job satisfaction, salary, time spent searching, and years of experience.

Hyperparameter Shuffling (Model-Choice Robustness)

```{r hyperparam-shuffling, eval=FALSE}
library(dplyr)
library(readr)

# Run the shuffling in main.r and load the output table
source("main.r")
shuffle <- read_csv("outputs/cart_hyperparam_shuffle.csv", show_col_types = FALSE)

best_row <- shuffle %>%
  filter(status == "ok") %>%
  arrange(xerror, desc(nsplit)) %>%
  slice(1)

best_row
```

This check keeps the imputed dataset fixed and varies cp, minsplit, and maxdepth to see how pruning and tree depth change the structure of the model. It is a model-choice robustness check, not a data-uncertainty check. The results are saved to [outputs/cart_hyperparam_shuffle.csv](outputs/cart_hyperparam_shuffle.csv).

In the shuffling, cp acts as the pruning strength and xerror is the cross-validated error from rpart. Higher cp values prune more aggressively and can remove all splits. This is visible in [outputs/cart_hyperparam_shuffle.csv](outputs/cart_hyperparam_shuffle.csv), where cp values of 0.008 to 0.02 yield nsplit = 0. The best-performing combination was cp = 0.001, minsplit = 20, and maxdepth = 5, with xerror = 0.9872315 and nsplit = 170. Under this setting, variable importance favored Years_Exp (1676.70) and Salary (872.31), with smaller contributions from Is_Remote (263.38) and Uses_AI (203.72). TimeSearching_Min did not appear in the main tree (importance = 0) for this specific imputation.

```{r}
# Placeholder: Hyperparameter shuffle xerror vs cp
```

This plot will show how pruning strength affects cross-validated error. A higher cp means stronger pruning and can collapse the tree into no splits. The lowest xerror with non-zero splits is the basis for selecting the final hyperparameters.

```{r}
# Placeholder: Variable importance bar chart for the main tree
```

The main-tree importance chart will highlight the primary drivers of job satisfaction under the best hyperparameter setting from the shuffling. It should emphasize that these importance values come from a single, fixed imputation and therefore reflect model-choice effects rather than imputation uncertainty.

Seed Sensitivity Check (Data Uncertainty)

```{r seed-sensitivity, eval=FALSE}
library(dplyr)
library(readr)

# Run the seed sensitivity check and load the summaries
source("cart_robustness_check.r")
seed_runs <- read_csv("outputs/cart_mice_seed_runs.csv", show_col_types = FALSE)
seed_averages <- read_csv("outputs/cart_mice_seed_averages.csv", show_col_types = FALSE)

seed_averages
```

This check fixes the model settings but re-imputes the missing data across several seeds. It uses seeds 42, 100, 200, 300, and 400 with m = 5 and maxit = 5, and it holds the CART settings constant at cp = 0.001, minsplit = 20, and maxdepth = 5. The full results are written to [outputs/cart_mice_seed_runs.csv](outputs/cart_mice_seed_runs.csv), and the per-seed averages are written to [outputs/cart_mice_seed_averages.csv](outputs/cart_mice_seed_averages.csv).

The seed-sensitivity check in [outputs/cart_mice_seed_averages.csv](outputs/cart_mice_seed_averages.csv) shows how results shift when the imputed data changes. Across seeds 42, 100, 200, 300, and 400, TimeSearching_Min becomes the dominant signal (range: 2444.99 to 2509.76). In the same runs, Years_Exp stays moderate (range: 970.66 to 1112.81) and Salary stays lower (range: 342.53 to 684.11). Uses_AI drops to near zero in most runs (range: 0 to 2.75), and Is_Remote stays very small (range: 1.23 to 4.56). The xerror values remain tightly clustered between 0.97838 and 0.97942, with split counts ranging from 159 to 192.

```{r}
# Placeholder: Seed sensitivity plot (importance by seed)
```

This plot will compare variable importance across imputation seeds to show how data uncertainty affects the model. Stable predictors will show consistent importance, while unstable predictors will fluctuate or drop out.

CP Profile (Pruning Check)

```{r cp-profile, eval=FALSE}
library(readr)

cp_profile <- read_csv("outputs/cart_cp_profile.csv", show_col_types = FALSE)
cp_profile
```

The CP profile is generated in [cart_robustness_check.r](cart_robustness_check.r) using a baseline imputation from the first seed. The profile table is saved to [outputs/cart_cp_profile.csv](outputs/cart_cp_profile.csv) and is used to explain how error changes as the tree is pruned. This gives a second, independent check on the pruning choice beyond the hyperparameter shuffling alone.

```{r}
# Placeholder: CP profile table or line plot
```

The CP profile summarizes the trade-off between tree size and cross-validated error. It helps justify why the selected tree complexity is not overfitting a single split of the data.

The Broad Trends vs. Granular Robustness

A key finding of this study comes from evaluating the data at two different levels of mathematical complexity. When the model is constrained to look for broad, macro-level trends, the time developers spend debugging or searching for answers does not predict their job satisfaction at all. At a career level, developers appear to experience a "productivity placebo." The rewarding feeling of moving faster with AI outweighs the reality of having a heavier workload later on (Becker et al., 2025).

However, we also stress-tested the model through rigorous cross-validation to ensure our findings were not just a fluke. By adjusting how the algorithm handles missing survey data and allowing the model to map more granular patterns, the picture changes. In these highly detailed models, the impact of simply using AI drops out entirely, while the time spent debugging emerges as a highly volatile and dominant predictor of satisfaction. This granular finding empirically validates the warnings from Afroz et al. (2025) about "spurious productivity." While high pay might keep developers satisfied overall, the hidden, unglamorous effort of reviewing and fixing AI-generated code remains a major pressure point when evaluating the micro-details of their work.

The Real Drivers of Satisfaction

Our analysis shows that AI tools are not the main drivers of developer well-being. Instead, years of experience and salary are the clear, dominant predictors of job satisfaction. This confirms that happiness at work is still tied to basic, foundational needs (Herzberg, 1959). Companies cannot simply introduce AI tools and expect morale to improve, especially since AI can sometimes drive developers into isolated silos, sacrificing team collaboration (Afroz et al., 2025). Instead, competitive pay remains the best way to keep staff. For example, the 56% wage premium for AI skills (PricewaterhouseCoopers, 2025) acts as "golden handcuffs" (Mitchell et al., 2001). This high pay keeps professionals satisfied overall, despite the daily friction of their AI-assisted workflows.

Furthermore, the model shows that using AI does not change job satisfaction for junior developers. Early-career professionals care much more about workplace flexibility, such as remote work (Dice, 2024). In contrast, AI acts as a small bonus for senior developers, specifically those earning over $102,000. This means GenAI is not a universal fix for low morale. It is simply a secondary perk for veterans who already have high pay, independence, and the experience needed to comfortably handle the technical debt that AI leaves behind.


```{r}
# Placeholder: Variable importance bar chart for the main tree
```


```{r}
# Placeholder: CP profile table or line plot
```

The CP profile summarizes the trade-off between tree size and cross-validated error. It helps justify why the selected tree complexity is not overfitting a single split of the data.

Interpretation

Taken together, the evidence suggests that the main tree is stable in its structure at the chosen hyperparameters, but the importance ranking is sensitive to imputation. When the tree is tuned for structure on a fixed imputation, the dominant drivers are years of experience and salary, and AI use appears only as a small, conditional factor. When the missing data is perturbed through different imputations, workflow friction rises to the top, while the direct effect of AI use becomes unstable.

This shows a weak and non-robust direct link between AI use and job satisfaction, while compensation, experience, and day-to-day workflow friction remain the strongest and most consistent signals.

## Chapter 5: Conclusion and Recommendations
Conclusion

Summarize key findings.

**Recommendations**

Provide practical suggestions

## References

Use APAformat.

## Appendices
Appendix A: Code
