---
title: "Culminating Activity"
author: "Aloba, Loyan Edrian | Dalmonte, Franklin | Delrio Martin Nicholas | Dorin, Homer Adriel | Rivera, Nythan"
date: ""

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Abstract
Provide a concise summary of the study including purpose, methods, results, and conclusions (150–250 words).

# Chapter 1: Introduction

## Background of the Study
The integration of Generative AI (GenAI) into software engineering has been aggressively marketed as an unprecedented paradigm shift in productivity. Industry narratives frequently promise that AI coding assistants will streamline workflows, accelerate output, and consequently elevate both financial compensation and developer well-being. However, emerging empirical research reveals a significant productivity trade-off inherent in GenAI adoption. Utilizing the SPACE framework for developer productivity, Afroz et al. (2025) found that while developers using AI tools experience localized spikes in activity and perceive themselves as operating faster, they do not necessarily produce better software or experience increased fulfillment. In fact, rigorous randomized controlled trials indicate that developers using early-2025 AI tools took 19% longer to complete complex tasks, despite subjectively believing they were working 20% faster—a cognitive disconnect establishing an industry-wide "productivity placebo" (Becker et al., 2025).
This paradox is driven by severe cognitive and operational friction. As AI accelerates initial code generation, it shifts systemic bottlenecks downstream, increasing pull request review times by 91% and inflating the volume of bugs per developer (Faros AI, 2025). Developers are forced to spend more time debugging hallucinated code and navigating complex, AI-generated architectural debt, which has been shown to actively damage macroscopic software delivery performance and system stability (DORA Research Program, 2024).
Consequently, this study investigates whether the financial rewards of AI adoption genuinely translate into higher job satisfaction, or if the persistent friction of modern coding creates a "GenAI Productivity-Satisfaction Tradeoff." In this paradigm, developers may be caught in a cycle where hyper-compensation—driven by a reported 56% AI skill premium (PricewaterhouseCoopers, 2025)—acts as a retention mechanism despite high stress, a concept previously identified in occupational psychology as job embeddedness or "golden handcuffs" (Mitchell et al., 2001). As AI increasingly shifts the developer's role from writing original code to reviewing and debugging AI output, the industry must question whether these tools actually improve the human experience of software engineering. Utilizing secondary data from the Stack Overflow 2024 Developer Survey (Stack Overflow, 2024), this study builds a predictive model to determine if AI tools genuinely improve job satisfaction, or if developers are simply compensated more to endure increased friction.


## Statement of the Problem
This study investigates the systemic impact of AI adoption on software developers. With Job Satisfaction as the primary target variable, the study asks:
To what extent does the use of AI development tools predict a developer’s overall job satisfaction?
How does workflow friction (time spent searching for answers and debugging) interact with AI usage to influence this satisfaction?
Does a higher salary offset the negative impact of workflow friction when predicting overall job satisfaction?


## Objectives of the Study
### General Objective
To investigate the "GenAI Productivity-Satisfaction Tradeoff" by developing a predictive model that evaluates how AI tool adoption, financial compensation, and workflow friction interact to determine a software developer's overall job satisfaction.

### Specific Objectives
To determine the extent to which AI tool adoption serves as a predictor for both financial compensation and workflow friction (time spent searching and debugging).

To identify the primary predictors of developer job satisfaction using a Classification and Regression Tree (CART) machine learning model.

To evaluate the interaction between compensation and workflow friction, specifically testing if a higher salary offsets the negative predictive impact of AI-induced friction on overall job satisfaction.


## Hypotheses (if applicable)
H0 (Null Hypothesis): AI usage, financial compensation, and workflow friction do not form a significant predictive model for developer job satisfaction, and no distinct tradeoff exists between these variables.

H1 (Alternative Hypothesis): The optimal predictive model for Job Satisfaction will demonstrate a significant tradeoff, where the positive predictive value of high financial compensation is canceled out by the negative predictive impact of AI-induced workflow friction.


## Significance of the Study
This research benefits technology companies, engineering managers, and developers. By identifying the gap between productivity metrics such as salary and output, and human metrics such as satisfaction, organizations can re-evaluate how they integrate AI. It shifts the focus from simply buying AI licenses to actively reducing the systemic friction that causes developer burnout.


## Scope and Limitations
The study analyzes secondary data from the Stack Overflow 2024 Developer Survey. The scope is strictly limited to respondents who identify as "Employed, full-time" to maintain a stable economic baseline.

A critical limitation is that survey data regarding compensation and satisfaction is highly susceptible to MNAR (Missing Not At Random) bias, as extreme earners and highly burned-out developers frequently skip these questions. To mitigate this, advanced algorithmic imputation using MICE via Classification and Regression Trees was utilized to predict and fill missing values rather than relying on listwise deletion.

This study does not attempt an exhaustive feature selection of the entire Stack Overflow dataset. Instead, the predictor variables (AI Usage, Salary, Time Spent Searching, Years of Experience, and Remote Status) were handpicked based on intuition and prior literature regarding developer productivity.  	


# Chapter 2: Review of Related Literature

## Related Studies
Summarize previous research.

## Related Literature


## Conceptual Framework
At the core of this framework is the ultimate target variable: Job Satisfaction.

Rather than utilizing isolated linear regressions, the framework employs a Decision Tree architecture where predictor variables interact organically:

Primary Predictor: AI Tool Adoption is positioned as the central independent variable, hypothesized to simultaneously impact financial compensation and workflow friction.

Interacting Predictors: Financial Compensation (Salary) and Workflow Friction (Time Spent Searching) act as mediating variables within the model.
Contextual Controls: Years of Experience and Remote Status serve as baseline variables to account for demographic and workplace variances.

The CART algorithm evaluates how these variables interact, specifically testing whether high compensation mathematically buffers the negative impact of AI-induced workflow friction when predicting final job satisfaction levels.

```{r conceptual-framework, echo=FALSE, fig.width=10, fig.height=5.8, out.width='100%', fig.cap="Conceptual Framework of the GenAI Productivity-Satisfaction Tradeoff"}
old_par <- par(mar = c(1, 1, 0.6, 1))

plot.new()
plot.window(xlim = c(0, 10), ylim = c(0, 8))
box(col = "#666666")
title(main = "Conceptual Framework of the GenAI Productivity-Satisfaction Tradeoff", cex.main = 1.0, line = 0.5)

draw_box <- function(xmin, ymin, xmax, ymax, label, col, border, cex = 0.88) {
  rect(xmin, ymin, xmax, ymax, col = col, border = border, lwd = 2)
Hyperparameter Shuffling (Model-Choice Robustness)
}

draw_arrow <- function(x0, y0, x1, y1, label = NULL, y_offset = 0) {
  arrows(x0, y0, x1, y1, length = 0.08, lwd = 1.8, col = "#666666")
  if (!is.null(label)) {
source("main.r")
shuffle <- read_csv("outputs/cart_hyperparam_shuffle.csv", show_col_types = FALSE)

best_row <- shuffle %>%
  filter(status == "ok") %>%
  arrange(xerror, desc(nsplit)) %>%
  slice(1)

best_row
```

This check keeps the imputed dataset fixed and varies cp, minsplit, and maxdepth to see how pruning and tree depth change the structure of the model. It is a model-choice robustness check, not a data-uncertainty check. The results are saved to [outputs/cart_hyperparam_shuffle.csv](outputs/cart_hyperparam_shuffle.csv).

Analysis pipeline used for this chapter:

1. [filter_columns.r](filter_columns.r) created [outputs/data_filtered_columns.csv](outputs/data_filtered_columns.csv).
2. [check_data.r](check_data.r) was used as an exploratory diagnostic pass on the raw and filtered data.
3. The imputation step was handled inside [main.r](main.r), where the MICE imputed dataset was created or reused from cache.
4. The CART model and the hyperparameter shuffling were then run on that imputed data.

The baseline seed was updated to 200 after the seed-sensitivity check showed that it had the lowest xerror among the tested seeds. That makes the main run easier to defend, because it uses the seed that performed best under the robustness check rather than the first seed chosen during the earlier trial run.

In the shuffling, cp acts as the pruning strength and xerror is the cross-validated error from rpart. Higher cp values prune more aggressively and can remove all splits. This is visible in [outputs/cart_hyperparam_shuffle.csv](outputs/cart_hyperparam_shuffle.csv), where cp values of 0.008 to 0.02 yield nsplit = 0. The best-performing combination was cp = 0.001, minsplit = 20, and maxdepth = 5, with xerror = 0.9872315 and nsplit = 170. Under the corrected baseline, TimeSearching_Min was the strongest predictor, followed by Years_Exp and Salary. Uses_AI and Is_Remote did not contribute meaningfully in that run.

```{r}
# Placeholder: Hyperparameter shuffle xerror vs cp
```

This plot will show how pruning strength affects cross-validated error. A higher cp means stronger pruning and can collapse the tree into no splits. The lowest xerror with non-zero splits is the basis for selecting the final hyperparameters.

```{r}
# Placeholder: Variable importance bar chart for the main tree
```

The main-tree importance chart will highlight the primary drivers of job satisfaction under the best hyperparameter setting from the shuffling. It should emphasize that these importance values come from a single, fixed imputation and therefore reflect model-choice effects rather than imputation uncertainty.

Seed Sensitivity Check (Data Uncertainty)

```{r seed-sensitivity, eval=FALSE}
library(dplyr)
library(readr)

# Run the seed sensitivity check and load the summaries
source("cart_robustness_check.r")
seed_runs <- read_csv("outputs/cart_mice_seed_runs.csv", show_col_types = FALSE)
seed_averages <- read_csv("outputs/cart_mice_seed_averages.csv", show_col_types = FALSE)

seed_averages
```

This check fixes the model settings but re-imputes the missing data across several seeds. It uses seeds 42, 100, 200, 300, and 400 with m = 5 and maxit = 5, and it holds the CART settings constant at cp = 0.001, minsplit = 20, and maxdepth = 5. The full results are written to [outputs/cart_mice_seed_runs.csv](outputs/cart_mice_seed_runs.csv), and the per-seed averages are written to [outputs/cart_mice_seed_averages.csv](outputs/cart_mice_seed_averages.csv).

The main run originally started with seed 42, but the sensitivity check showed that seed 200 gave the best overall xerror. That is why the baseline was updated to seed 200 for the final main run.

The seed-sensitivity check in [outputs/cart_mice_seed_averages.csv](outputs/cart_mice_seed_averages.csv) shows how results shift when the imputed data changes. Across seeds 42, 100, 200, 300, and 400, TimeSearching_Min becomes the dominant signal (range: 2444.99 to 2509.76). In the same runs, Years_Exp stays moderate (range: 970.66 to 1112.81) and Salary stays lower (range: 342.53 to 684.11). Uses_AI drops to near zero in most runs (range: 0 to 2.75), and Is_Remote stays very small (range: 1.23 to 4.56). The xerror values remain tightly clustered between 0.97838 and 0.97942, with split counts ranging from 159 to 192.

```{r}
# Placeholder: Seed sensitivity plot (importance by seed)
```

This plot will compare variable importance across imputation seeds to show how data uncertainty affects the model. Stable predictors will show consistent importance, while unstable predictors will fluctuate or drop out.

CP Profile (Pruning Check)

```{r cp-profile, eval=FALSE}
library(readr)

cp_profile <- read_csv("outputs/cart_cp_profile.csv", show_col_types = FALSE)
cp_profile
```

The CP profile is generated in [cart_robustness_check.r](cart_robustness_check.r) using a baseline imputation from the first seed. The profile table is saved to [outputs/cart_cp_profile.csv](outputs/cart_cp_profile.csv) and is used to explain how error changes as the tree is pruned. This gives a second, independent check on the pruning choice beyond the hyperparameter shuffling alone.
draw_arrow(6.1, 5.8, 7.2, 4.2, "Higher compensation may\nreduce the negative effect\nof friction", 0.12)
draw_arrow(6.1, 2.5, 7.2, 3.7, "Workflow friction may\nlower job satisfaction", -0.1)
draw_arrow(3.0, 6.8, 4.0, 6.1)
draw_arrow(8.2, 1.6, 8.1, 3.0)

par(old_par)
```

The diagram illustrates the interacting variables within the predictive CART model. AI Tool Adoption acts as the primary predictor, hypothesized to trigger a dual pathway: an increase in Financial Compensation and a simultaneous increase in Workflow Friction. These mediating variables converge on the target variable, Job Satisfaction, where the algorithm will test if the financial benefits successfully buffer the friction-induced burnout. In this statistical context, "buffering" refers to a moderating effect where high compensation offsets, mitigates, or increases a developer's tolerance for the negative psychological toll of AI-driven workflow friction (Lam, et al., 2025). Finally, Prior Years of Experience and Remote Work Status are integrated into the system as contextual controls that adjust the model's baseline predictions. Unlike a sequential linear regression, the CART algorithm evaluates how all of these variables split and compound simultaneously to determine the final satisfaction outcome.

## Chapter 3: Methodology
Research Design

This study employs a quantitative, descriptive-correlational research design. It utilizes multiple linear regression modeling to examine the predictive relationships between AI adoption, compensation, and workflow friction.

Data Collection


```{r}
# Load dataset
# data <- read.csv("your_data.csv")
# head(data)
```

Variables Description


Data Analysis Techniques


## Chapter 4: Results and Discussion
Descriptive Statistics

```{r descriptive-statistics}
library(dplyr)
library(readr)

df <- read_csv("outputs/data_imputed_primary.csv", show_col_types = FALSE)

summary_df <- df %>%
  summarise(
    n = n(),
    Job_Satisfaction_mean = mean(Job_Satisfaction, na.rm = TRUE),
    Job_Satisfaction_sd = sd(Job_Satisfaction, na.rm = TRUE),
    Salary_mean = mean(Salary, na.rm = TRUE),
    Salary_sd = sd(Salary, na.rm = TRUE),
    TimeSearching_Min_mean = mean(TimeSearching_Min, na.rm = TRUE),
    TimeSearching_Min_sd = sd(TimeSearching_Min, na.rm = TRUE),
    Years_Exp_mean = mean(Years_Exp, na.rm = TRUE),
    Years_Exp_sd = sd(Years_Exp, na.rm = TRUE),
    Uses_AI_rate = mean(as.numeric(as.character(Uses_AI)) == 1, na.rm = TRUE),
    Is_Remote_rate = mean(as.numeric(as.character(Is_Remote)) == 1, na.rm = TRUE)
  )

summary_df
```

Visualization
```{r}
# Placeholder for charts/plots
```


This section summarizes the sample used in the analysis, including central tendencies for job satisfaction, salary, time spent searching, and years of experience.

Hyperparameter Shuffling (Model-Choice Robustness)

```{r hyperparam-shuffling, eval=FALSE}
library(dplyr)
library(readr)

# Run the shuffling in main.r and load the output table
source("main.r")
shuffle <- read_csv("outputs/cart_hyperparam_shuffle.csv", show_col_types = FALSE)

best_row <- shuffle %>%
  filter(status == "ok") %>%
  arrange(xerror, desc(nsplit)) %>%
  slice(1)

best_row
```

This check keeps the imputed dataset fixed and varies cp, minsplit, and maxdepth to see how pruning and tree depth change the structure of the model. It is a model-choice robustness check, not a data-uncertainty check. The results are saved to [outputs/cart_hyperparam_shuffle.csv](outputs/cart_hyperparam_shuffle.csv).

In the shuffling, cp acts as the pruning strength and xerror is the cross-validated error from rpart. Higher cp values prune more aggressively and can remove all splits. This is visible in [outputs/cart_hyperparam_shuffle.csv](outputs/cart_hyperparam_shuffle.csv), where cp values of 0.008 to 0.02 yield nsplit = 0. The best-performing combination was cp = 0.001, minsplit = 20, and maxdepth = 5, with xerror = 0.9872315 and nsplit = 170. Under the corrected baseline, TimeSearching_Min was the strongest predictor, followed by Years_Exp and Salary. Uses_AI and Is_Remote did not contribute meaningfully in that run.

```{r}
# Placeholder: Hyperparameter shuffle xerror vs cp
```

This plot will show how pruning strength affects cross-validated error. A higher cp means stronger pruning and can collapse the tree into no splits. The lowest xerror with non-zero splits is the basis for selecting the final hyperparameters.

```{r}
# Placeholder: Variable importance bar chart for the main tree
```

The main-tree importance chart will highlight the primary drivers of job satisfaction under the best hyperparameter setting from the shuffling. It should emphasize that these importance values come from a single, fixed imputation and therefore reflect model-choice effects rather than imputation uncertainty.

Seed Sensitivity Check (Data Uncertainty)

```{r seed-sensitivity, eval=FALSE}
library(dplyr)
library(readr)

# Run the seed sensitivity check and load the summaries
source("cart_robustness_check.r")
seed_runs <- read_csv("outputs/cart_mice_seed_runs.csv", show_col_types = FALSE)
seed_averages <- read_csv("outputs/cart_mice_seed_averages.csv", show_col_types = FALSE)

seed_averages
```

This check fixes the model settings but re-imputes the missing data across several seeds. It uses seeds 42, 100, 200, 300, and 400 with m = 5 and maxit = 5, and it holds the CART settings constant at cp = 0.001, minsplit = 20, and maxdepth = 5. The full results are written to [outputs/cart_mice_seed_runs.csv](outputs/cart_mice_seed_runs.csv), and the per-seed averages are written to [outputs/cart_mice_seed_averages.csv](outputs/cart_mice_seed_averages.csv).

The seed-sensitivity check in [outputs/cart_mice_seed_averages.csv](outputs/cart_mice_seed_averages.csv) shows how results shift when the imputed data changes. Across seeds 42, 100, 200, 300, and 400, TimeSearching_Min becomes the dominant signal (range: 2444.99 to 2509.76). In the same runs, Years_Exp stays moderate (range: 970.66 to 1112.81) and Salary stays lower (range: 342.53 to 684.11). Uses_AI drops to near zero in most runs (range: 0 to 2.75), and Is_Remote stays very small (range: 1.23 to 4.56). The xerror values remain tightly clustered between 0.97838 and 0.97942, with split counts ranging from 159 to 192.

```{r}
# Placeholder: Seed sensitivity plot (importance by seed)
```

This plot will compare variable importance across imputation seeds to show how data uncertainty affects the model. Stable predictors will show consistent importance, while unstable predictors will fluctuate or drop out.

CP Profile (Pruning Check)

```{r cp-profile, eval=FALSE}
library(readr)

cp_profile <- read_csv("outputs/cart_cp_profile.csv", show_col_types = FALSE)
cp_profile
```

The CP profile is generated in [cart_robustness_check.r](cart_robustness_check.r) using a baseline imputation from the first seed. The profile table is saved to [outputs/cart_cp_profile.csv](outputs/cart_cp_profile.csv) and is used to explain how error changes as the tree is pruned. This gives a second, independent check on the pruning choice beyond the hyperparameter shuffling alone.

```{r}
# Placeholder: CP profile table or line plot
```

The CP profile summarizes the trade-off between tree size and cross-validated error. It helps justify why the selected tree complexity is not overfitting a single split of the data.

The Robustness Story

The corrected baseline does not support AI usage as a strong driver of job satisfaction. In the main tree, TimeSearching_Min is the dominant split variable, with Years_Exp and Salary following behind. This means the model is showing that workflow friction is the strongest signal in the fixed-imputation run.

The robustness checks explain why this matters. When the model settings are shuffled, the same pattern stays visible: the chosen tree still favors TimeSearching_Min under the best-performing settings, while AI usage remains weak and unstable. When the missing data is perturbed through the seed sensitivity check, the same broad conclusion holds. TimeSearching_Min stays dominant, Years_Exp and Salary remain secondary, and Uses_AI contributes little or nothing in most runs.

The Real Drivers of Satisfaction

The updated results show that job satisfaction is shaped most strongly by workflow friction, years of experience, and salary. This is closer to Herzberg's (1959) view that satisfaction rests on basic working conditions and compensation, not on a single new tool. AI may still matter in limited cases, but it is not the main driver of the tree. In the corrected baseline, the model points first to the time spent searching and debugging, then to experience and pay.

This means GenAI is not a universal fix for low morale. It behaves more like a conditional feature than a core predictor. The 56% wage premium for AI skills (PricewaterhouseCoopers, 2025) can still work like "golden handcuffs" (Mitchell et al., 2001), but the tree does not show AI usage itself as the central force shaping satisfaction.

Interpretation

Taken together, the evidence suggests that the main tree is stable in its structure at the chosen hyperparameters, but the importance ranking is sensitive to imputation. In the corrected baseline, TimeSearching_Min is the strongest predictor, with Years_Exp and Salary trailing behind. When the missing data is perturbed through different imputations, workflow friction remains the top signal, while Uses_AI stays weak and unstable.

This shows a weak and non-robust direct link between AI use and job satisfaction, while workflow friction, years of experience, and salary remain the strongest and most consistent signals.

## Chapter 5: Conclusion and Recommendations
Conclusion

This research investigated the systemic impact of AI adoption on software developers, specifically examining whether workflow friction and compensation help explain job satisfaction. The findings show that the direct link between AI tool adoption and overall job satisfaction is weak. In the corrected baseline, the main driver is TimeSearching_Min, followed by Years_Exp and Salary.
The seed-sensitivity results reinforce the same direction. When the missing data is re-imputed under different seeds, TimeSearching_Min stays dominant, while AI usage remains small or drops out. This means the data does not support a story where AI itself becomes a major driver once the tree is tuned more deeply. Instead, the model keeps pointing back to the daily effort of searching, debugging, and managing workflow friction.
Regarding the question of whether salary offsets this friction, the evidence suggests a practical rather than a mathematical tradeoff. Higher pay still matters, but it does not erase the effect of workflow friction. It simply helps explain why developers remain satisfied enough overall despite the strain of their work. These findings align with Afroz et al. (2025) and the broader warning about "spurious productivity": surface-level acceleration does not necessarily mean better work experience.
Consequently, the alternative hypothesis (H1) is rejected. The expected tradeoff does not appear as a direct cancellation effect in the tree. Overall job satisfaction is driven most strongly by workflow friction, years of experience, and salary, while AI adoption remains a weak and conditional factor.


Summarize key findings.

**Recommendations**

Provide practical suggestions

## References

Use APAformat.

## Appendices
Appendix A: Code
