---
title: "The GenAI Productivity-Satisfaction Tradeoff"
author: "G2G7 - Aloba, Del Monte, Del Rio, Dorin, Rivera"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: lumen
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```

## Executive Summary

This Exploratory Data Analysis (EDA) visualizes data that will be used to measure the tradeoff between salary and job satisfaction associated with Generative AI adoption among full-time software developers, using the Stack Overflow 2024 survey.

The dataset is filtered to respondents with `Employment == "Employed, full-time"` and the following analysis variables: `Salary`, `Uses_AI`, `Years_Exp`, `Job_Satisfaction`, `Is_Remote`, and `TimeSearching`.

The analysis focuses on three simple comparisons: salary by AI adoption, job satisfaction by time spent searching/debugging, and salary trajectories by experience and AI adoption.

```{r load-data}
library(dplyr)
library(ggplot2)
library(readr)
library(scales)

df <- read_csv("outputs/data_imputed.csv", show_col_types = FALSE)

df_viz <- df %>%
  mutate(
    AI_Status = factor(ifelse(Uses_AI == 1, "Adopts AI", "No AI"), levels = c("No AI", "Adopts AI")),
    Remote_Status = factor(ifelse(Is_Remote == 1, "Remote", "Hybrid / In-person"), levels = c("Hybrid / In-person", "Remote")),
    TimeSearching = as.factor(TimeSearching)
  )

cat("Observations:", nrow(df_viz), "full-time developers\n")
cat("Columns used:", paste(names(df_viz), collapse = ", "), "\n")
```

## 1. Financial Reality: Salary by AI Adoption

Does AI adoption associate with higher reported salary?

```{r wage-gap, fig.height=4}
ggplot(df_viz, aes(x = AI_Status, y = Salary, fill = AI_Status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  scale_fill_manual(values = c("No AI" = "#bdc3c7", "Adopts AI" = "#3498db")) +
  scale_y_continuous(labels = dollar_format()) +
  coord_cartesian(ylim = c(0, 250000)) +
  theme_minimal(base_size = 12) +
  labs(title = "Salary Distribution by AI Adoption", x = NULL, y = "Annual Salary (USD)") +
  theme(legend.position = "none")
```

## 2. Job Satisfaction vs. Time Spent Searching

Does more time spent searching/debugging associate with lower job satisfaction?

```{r friction-penalty, fig.height=4}
ggplot(df_viz, aes(x = TimeSearching, y = Job_Satisfaction, fill = TimeSearching)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal(base_size = 12) +
  labs(title = "Job Satisfaction by Time Spent Searching/Debugging", x = "Time Searching (categorical)", y = "Job Satisfaction (0-10)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

## 3. Salary Trajectory by Experience and AI Adoption

Does AI adoption associate with different salary trajectories across years of experience?

```{r experience-multiplier, fig.height=4}
ggplot(df_viz, aes(x = Years_Exp, y = Salary, color = AI_Status)) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  scale_color_manual(values = c("No AI" = "#7f8c8d", "Adopts AI" = "#2980b9")) +
  scale_y_continuous(labels = dollar_format()) +
  coord_cartesian(ylim = c(0, 250000), xlim = c(0, 35)) +
  theme_minimal(base_size = 12) +
  labs(title = "Salary by Experience and AI Adoption", x = "Years of Experience", y = "Annual Salary (USD)") +
  theme(legend.position = "bottom")
```

## 4. Does Remote Work Influence?

Because `Is_Remote` is a filtered feature, this summarizes how AI adoption appears across remote vs hybrid/in-person work modes.

```{r remote-context, fig.height=4}
ggplot(df_viz, aes(x = Remote_Status, fill = AI_Status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("No AI" = "#7f8c8d", "Adopts AI" = "#2980b9")) +
  theme_minimal(base_size = 12) +
  labs(
    title = "AI Adoption Share by Work Mode",
    x = "Work Mode",
    y = "Share of Respondents",
    fill = "AI Status"
  )
```


## 5. Does AI actually help with problem solving?

Does AI adoption mitigate the pain of workflow friction?

```{r friction-buffer, fig.height=5, fig.width=10}
ggplot(df_viz, aes(x = TimeSearching, y = Job_Satisfaction, fill = AI_Status)) +
  geom_boxplot(alpha = 0.8, position = position_dodge(width = 0.8), color = "#2c3e50") +
  scale_fill_manual(values = c("No AI" = "#bdc3c7", "Adopts AI" = "#3498db")) +
  theme_minimal(base_size = 12) +
  labs(
    title = "AI & Coding Friction", 
    subtitle = "Job Satisfaction across Searching Time based on AI use  ",
    x = "Daily Time Spent Searching for Answers", 
    y = "Job Satisfaction Score (0-10)",
    fill = "Workflow Context"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"), 
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```


## 6. AI Use Years-Experience Adoption

Is AI just a junior developer tool, or is it industry-wide?
```{r adoption-density, fig.height=4, fig.width=9}
ggplot(df_viz, aes(x = Years_Exp, fill = AI_Status)) +
  geom_density(alpha = 0.6, color = "#2c3e50") +
  scale_fill_manual(values = c("No AI" = "#bdc3c7", "Adopts AI" = "#3498db")) +
  coord_cartesian(xlim = c(0, 35)) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Adoption Curve", 
    subtitle = "Density of AI Usage Across Professional Seniority",
    x = "Years of Professional Experience", 
    y = "Density of Developers",
    fill = "AI or No AI?"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```